---
title: "MetNet - workshop Toulouse"
author: "Thomas Naake"
date: "9/23/2021"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: TRUE
    theme: united
    number_sections: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("igraph"))
knitr::opts_knit$set(root.dir = "~/Presentation/20210923_Toulouse/")
```

# Prepare the environment and load the data

Load, if not done yet, the `dplyr` and `igraph` package first.
```{r prepare_environment, message=FALSE, warning=FALSE}
library(MetNet)

## print the version of MetNet
package.version("MetNet")
```

data requirements:
- $m~\times~n$ matrix, where columns denote the different samples $n$ and where
  $m$ features are present,
- proper data normalization and transformation,
- information on m/z values has to be stored in a vector of length 
  $\vert m \vert$ in the column `"mz"`,
- optionally, information on retention time in the column `"rt"`.

```{r laod_data}
peaklist_vals <- openxlsx::read.xlsx(
    xlsxFile = "20210219_CelegansPathogenMicrobiome_RP_pos_Cluster_char.xlsx",
    sheet = "Stage 3", startRow = 1, colNames = TRUE, rowNames = TRUE,
    skipEmptyRows = FALSE, skipEmptyCols = FALSE, na.strings = "")

## remove the superfluous columns in peaklist_vals (only keep samples)
cols_vals <- which(colnames(peaklist_vals) == "QC_cond_2-A,2_10_29503")
peaklist_vals <- peaklist_vals[, 1:cols_vals]

## extract meta data and intensities from peaklist_vals
peaklist_meta <- t(peaklist_vals[1:10, ]) %>% 
    as.data.frame()
colnames(peaklist_meta) <- make.names(colnames(peaklist_meta))
peaklist_vals <- peaklist_vals[11:nrow(peaklist_vals), ] %>% 
    as.matrix()
mode(peaklist_vals) <- "numeric"

## add the information on the features (mz, retention time)
peaklist_feat <- read.table(file = "Experimental_Data.csv", sep = ",", header = TRUE)
rownames(peaklist_feat) <- peaklist_feat[, "Name"]
```

## Quality check with MatrixQCvis

```{r matrixqcvis, eval = 1:5, message=FALSE}
## create SummarizedExperiment object
library(SummarizedExperiment)
se <- SummarizedExperiment(assays = peaklist_vals, rowData = peaklist_feat, 
                     colData = peaklist_meta)

library(MatrixQCvis)
shinyQC(se)
```

Do the normalization, batch correction, transformation after removing the blank.
```{r, message=FALSE}
se <- se[, !grepl(colnames(se), pattern = "blank")]

## remove low-abundant features from se (remove the features with more than
## 30 missing values)
se <- se[!apply(assay(se), 1, function(x) sum(is.na(x))) > 30, ]

## normalization
peaklist_vals <- assay(se) %>%
    normalizeAssay(., method = "quantile division", probs = 0.75)

## batch correction
peaklist_vals_b <- peaklist_vals %>% 
    imputeAssay(., method = "MinDet") %>%
    sva::ComBat(., batch = se$Batch)
peaklist_vals_b[is.na(peaklist_vals)] <- NA

## check the variance homogeneity and perform vsn transformation
vsn::meanSdPlot(peaklist_vals_b)
peaklist_vals <- vsn::vsn2(peaklist_vals_b)
vsn::meanSdPlot(peaklist_vals)

## create a peaklist object that contains information on the mz and rt and 
## the transformed intensities
peaklist <- data.frame(mz = rowData(se)[, "mz"], rt = rowData(se)[, "RT"], 
                  peaklist_vals@hx)

## assign new rownames to peaklist (composed of mz and RT information)
rn <- paste0("m", round(peaklist[, "mz"], 3), "_", round(peaklist[, "rt"], 2))
peaklist <- peaklist[!duplicated(rn), ]
rownames(peaklist) <- rn[!duplicated(rn)]

## remove the columns with the QCs
peaklist <- peaklist[, !grepl(colnames(peaklist), pattern = "QC_")]
```

# Structural AdjacencyMatrix object

```{r define_transformations}
transformations <- rbind(
    c("Hydroxylation (-H)", "O", 15.9949146221, "-"),
    c("Malonyl group (-H2O)", "C3H2O3", 86.0003939305, "+"),
    c("Monosaccharide (-H2O)", "C6H10O5", 162.0528234315, "-"))

## convert to data frame
transformations <- data.frame(
    group = transformations[, 1],
    formula = transformations[, 2],
    mass = as.numeric(transformations[, 3]),
    rt = transformations[, 4])
```

Create the structural adjacency matrix by running the function `structural`.
```{r create_struct_adj}
struct_adj <- structural(x = peaklist, transformation = transformations,
                         ppm = 10)
```

Print information on the object `struct_adj`.
```{r struct_adj_information}
struct_adj

## get information on type, directed, thresholded
struct_adj@type
struct_adj@directed
struct_adj@thresholded

## show the stored adjacency matrices
assay(struct_adj, "binary")[26:28, 614:616]
assay(struct_adj, "transformation")[26:28, 614:616]
assay(struct_adj, "mass_difference")[26:28, 614:616]

## how many edges are there in total
assay(struct_adj, "binary") %>% 
    table()
```

## Refining the structural AdjacencyMatrix object (rtCorrection)

Use the retention time information to remove false positive associations
by accounting for the polarity change. Use the function `rtCorrection`. The
slot will change from `FALSE` (`struct_adj`) to `TRUE` (`struct_adj_rt`).

```{r create_struct_adj_rt}
struct_adj_rt <- rtCorrection(am = struct_adj, x = peaklist, 
                              transformation = transformations)
struct_adj
struct_adj_rt

## how many edges are there in total
assay(struct_adj_rt, "binary") %>% 
    table()
```

Get information which features link to each other
```{r struct_df}
## show the information in struct_adj_rt in tabular format
struct_df <- as.data.frame(struct_adj_rt)
filter(struct_df, binary == 1) %>% 
    head()
```

Mass-difference distribution of the data. The number of determined
mass differences can be displayed by using the `mz_vis` function 
```{r mz_summary}
mz_sum <- mz_summary(struct_adj_rt)
mz_sum

## visualize the numbers
mz_vis(mz_sum)
```

## Creating a directed structural AdjacencyMatrix object 

```{r create_directed}
## undirected
struct_adj_undir <- structural(x = peaklist,
        transformation = transformations, ppm = 10, directed = FALSE) %>%
    rtCorrection(am = ., x = peaklist, transformation = transformations)
struct_adj_undir

## directed
struct_adj_dir <- structural(x = peaklist, 
        transformation = transformations, ppm = 10, directed = TRUE) %>%
    rtCorrection(am = ., x = peaklist, transformation = transformations)
struct_adj_dir

## plot the undirected matrix
adj <- assay(struct_adj_undir, "binary")
g_undirected_comp <- graph_from_adjacency_matrix(adjmatrix = adj, 
        mode = "directed", weighted = NULL) %>%
    components()
inds <- g_undirected_comp$membership %in% which(g_undirected_comp$csize > 10)

## create a graph from the cut adjacency matrix and plot the undirected graph
g_undirected <- graph_from_adjacency_matrix(adj[inds, inds], mode = "directed")
plot(g_undirected, edge.width = 1, edge.arrow.size = 0.3,
     vertex.label.cex = 0.4, vertex.size = 0.5, edge.color = "grey")

## plot the directed matrix
adj <- assay(struct_adj_dir, "binary")
g_directed_comp <- graph_from_adjacency_matrix(adjmatrix = adj, 
          mode = "directed", weighted = NULL) %>%
      components()
inds <- g_directed_comp$membership %in% which(g_directed_comp$csize > 10)

## create a graph from the cut adjacency matrix and plot the directed graph
g_directed <- graph_from_adjacency_matrix(adj[inds, inds], mode = "directed")
plot(g_directed, edge.width = 1, edge.arrow.size = 0.3,
     vertex.label.cex = 0.4, vertex.size = 0.5, edge.color = "grey")
```


## Visualize the structural AdjacencyMatrix

Use the `igraph` package to calculate a graph from the adjacency matrix. 
Only show those components with more than 10 members.

```{r vis_struct_adj, eval = FALSE, echo = FALSE}
adj <- assay(struct_adj_rt, "binary")
g_comp <-  graph_from_adjacency_matrix(adj, mode = "undirected") %>% 
    components()
inds <- g_comp$membership %in% which(g_comp$csize > 10)
g <- graph_from_adjacency_matrix(adj[inds, inds], mode = "undirected")

## finally plot the graph
plot(g, edge.width = 1, edge.arrow.size = 0.3, vertex.label = NA, 
    vertex.size = 0.5, edge.color = "grey")
```

# Statistical AdjacencyMatrix object

Create the adjacency matrix based on Pearson and Spearman correlation.
```{r create_stat_adj, warning=FALSE}
## the function takes as input a numerical matrix with 
## (transformed) intensities
stat_adj <- select(peaklist, -c("mz", "rt")) %>% 
    as.matrix() %>%
    statistical(., model = c("pearson", "spearman"), p.adjust = "BH")

## print information on stat_adj
stat_adj
```

Return the edges with Pearson correlation coefficients $> 0.6$ and 
Spearman correlation coefficients $> 0.6$ and adjusted p-values $< 0.05$.
```{r create_stat_adj_thr}
## define the filter criteria
args_thr <- list(
    filter = "abs(pearson_coef) > 0.6 & abs(spearman_coef) > 0.6 & pearson_pvalue < 0.05 & spearman_pvalue < 0.05")

## apply the filter criteria on stat_adj and add an assay containing
## binary information
stat_adj_thr <- threshold(am = stat_adj, type = "threshold", 
                          args = args_thr)
stat_adj_thr

## show the information in stat_adj_thr in tabular format
stat_adj_thr_df <- as.data.frame(stat_adj_thr)
head(stat_adj_thr_df)
```

# Combine the structural and statistical AdjacencyMatrix object

```{r create_comb_adj}
comb_adj <- combine(am_structural = struct_adj_rt, 
                    am_statistical = stat_adj_thr)

## print information on stat_adj
comb_adj

## show the information in comb_adj in tabular format
comb_adj_df <- as.data.frame(comb_adj)
filter(comb_adj_df, combine_binary == 1) %>% 
    head()
```

## Visualization

Visualize the final adjacency matrix that combines the information from 
both structural and statistical information.
```{r vis_comb_adj}
adj <- assay(comb_adj, "combine_binary")
g_comp <- graph_from_adjacency_matrix(adj, mode = "undirected") %>%
    components()
inds <- g_comp$membership %in% which(g_comp$csize > 3)

## create a graph from the cut adjacency matrix
g <- graph_from_adjacency_matrix(adj[inds, inds], mode = "undirected")

## finally plot the graph
plot(g, edge.width = 1, vertex.label.cex = 0.5, vertex.size = 0.5,
     edge.color = "grey")
```


```{r save_objects, echo = FALSE, eval = TRUE}
save(list = ls(), file = "MetNet_workshop_Celegans_objects.RData")
```